Value Ordering, Discrepancies, and Restarts in Subgraph Solvers: \\ Some Half-Baked Mucking Around

### 2017-09-26 paper/paper.tex
More thingies

\title{Value-Ordering, Discrepancies, and Restarts in Subgraph Solvers: \\ Some Half-Baked Mucking
Around}
\author{Ciaran McCreesh\thanks{This work was supported by the Engineering and Physical Sciences
    Research Council [grant number EP/026842/1]}\\ University of Glasgow, Glasgow, Scotland \\
    ciaran.mccreesh@glasgow.ac.uk}

@@

\begin{abstract}
    Modern subgraph isomorphism solvers use reasonably good value-ordering heuristics to direct
    search, but they are not perfect, particularly for early choices. We investigate two mechanisms
    for offsetting this weakness: discrepancy searches, and restarts. Our results suggest that
    discrepancy search is too expensive to be effective, but that restarts with a random
    value-ordering heuristic beat tailored heuristics. We then design a slightly-random
    value-ordering heuristic which, when combined with restarts, performs ?? times better on
    satisfiable instances without substantially weakening performance on unsatisfiable instances.
\end{abstract}

@@

The current strongest subgraph isomorphism solver uses degree as a value-ordering heuristic. This is
much better than not having a value-ordering heuristic, but it is far from perfect, particularly for
early branching choices. There are two reasons we suspect we can do better. Firstly, . Secondly, even if degree
were a perfect source of information, many graphs do not have a large degree distribution, and so we
still need to determine \emph{which} vertex of highest degree would be best.

We could try to introduce tiebreaking mechanisms. However, given the extreme cost of making an
incorrect branching choice early on in search, we believe it is more fruitful to investigate
alternatives to simple backtracking (depth-first) search.

@@

?? Started with the Glasgow algorithm, removed parts which did not contribute hugely to speedups,
and ended up with a simpler algorithm that performs as well in practice. In particular, large
graphs, no ILF, and not doing supplemental graphs of length 3.

\newcommand{\siplabel}[1]{\label{line:sip:#1}}
\begin{algorithm}[p]\DontPrintSemicolon\small
    \begin{tikzpicture}[remember picture,overlay]
        \coordinate (restart1sc) at ($(pic cs:restart1s) + (0, 0.15)$);
        \coordinate (restart1ec) at ($(pic cs:restart1e) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart1sc) (restart1ec)] { };
        \coordinate (restart2sc) at ($(pic cs:restart2s) + (0, 0.15)$);
        \coordinate (restart2ec) at ($(pic cs:restart2e) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart2sc) (restart2ec)] { };
        \coordinate (restart3sc) at ($(pic cs:restart3s) + (0, 0.15)$);
        \coordinate (restart3ec) at ($(pic cs:restart3e) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart3sc) (restart3ec)] { };
        \coordinate (restart4sc) at ($(pic cs:restart4s) + (0, 0.15)$);
        \coordinate (restart4ec) at ($(pic cs:restart4e) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart4sc) (restart4ec)] { };
        \coordinate (restart5sc) at ($(pic cs:restart5s) + (0, 0.15)$);
        \coordinate (restart5ec) at ($(pic cs:restart5e) - (0, 0.10)$);
        \coordinate (restart5rc) at ($(pic cs:restart5r) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart5sc) (restart5ec) (restart5rc)] { };
        \coordinate (restart6sc) at ($(pic cs:restart6s) + (0, 0.15)$);
        \coordinate (restart6ec) at ($(pic cs:restart6e) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart6sc) (restart6ec)] { };
        \coordinate (restart7sc) at ($(pic cs:restart7s) + (0, 0.15)$);
        \coordinate (restart7ec) at ($(pic cs:restart7e) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart7sc) (restart7ec)] { };
        \coordinate (restart8sc) at ($(pic cs:restart8s) + (0.05, 0.15)$);
        \coordinate (restart8ec) at ($(pic cs:restart8e) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart8sc) (restart8ec)] { };
    \end{tikzpicture}

@@

        \nl \tikzmark{restart2s}Trigger nogoods containing only one clause\tikzmark{restart2e} \;

@@

\smallskip

@@

    \nl $D_v \gets \textnormal{a domain from}~D~\textnormal{chosen by variable heuristic}$ \siplabel{pickd} \;

@@

    \nl \ForEach {$v' \in D_v~\textnormal{ordered by chosen value heuristic}$\siplabel{parforeach}}{

@@

            \nl $S \gets \FuncSty{search}(\{ E \in D' : \left|E\right| > 1 \},$ \\\hspace*{3em} $B
            \wedge (v \mapsto v'))$ \;

@@

                \nl Post nogoods $\{B \wedge (v \mapsto c) \Rightarrow \bot : c \in C \}$\tikzmark{restart5r}\;

@@

        \nl \tikzmark{restart7s}Trigger nogoods containing $(v \mapsto v')$\tikzmark{restart7e} \;

@@

\caption{Subgraph isomorphism. Highlighted lines are ignored for the non-restarting algorithm.}

@@

?? Variable ordering is smallest domain first with degree tiebreaking. This is not the focus of this
paper.

In our experiments, we compare five value-ordering heuristics.

\begin{description}
    \item[Tailored] Where we select vertices from highest degree to lowest degree (which is done by
        permuting the graph at the top of search).
    \item[Anti] Where we select vertices from lowest degree to highest degree.
    \item[Tiebreaking] Where we select vertices from highest degree to lowest degree,
        tiebreaking on highest sum of neighbour degrees.
    \item[Random] Where we select a branching vertex at random by (dynamically, that is, we shuffle
        the vertices before entering the for loop, rather than by randomly permuting the graph once at the
        top of search).
    \item[Weighted] Where we shuffle the vertices before each for loop, but instead of
        picking a vertex uniformly at random, we use a weight of
        $2^{\left|\neighbourhood(v)\right|}$ when deciding whether to branch on vertex $v$
        next.\footnote{Actually, we cheat a bit to avoid range issues with random number generation:
        we give the vertex with highest degree in the entire graph weight $2^{50}$, and then give
        lower degree vertices smaller powers, giving every vertex weight at least 1. This is a bit
        of hack but getting it perfectly right is hard.} This scheme heavily prefer vertices of high
        degree, but will give the same chance of being selected next to two vertices of equal
        degree. This also introduces an element of randomness, which will be important for restarts.
\end{description}

@@

?? Not dealing with binary search. Make it binary. Other versions where we increase the discrepancy
each time the degree changes, and where we count every ``against'' branch as one discrepancy.

@@

?? Aimed primarily at value-ordering, not value-ordering.

?? Value-ordering options

?? Propagation

?? We use the Luby scheme, because everyone else does. We count the number of backtracks (that is,
when we reach the end of the main for loop) to decide when to restart. Following established wisdom,
we multiply each item in the sequence by a magic constant. Preliminary experiments demonstrated that
this is useful, but we failed to determine a principled way of selecting the constant's value, so we
fell back on divine revelation and set it to 666. As with everyone else, we are not entirely clear
why we are doing this: it could be to allow the solver to spend more time deep in search, it could
be to reduce the overheads from repeatedly visiting inner nodes, or it could be to keep the nogoods
small.

?? Increasing nogoods etc. We are not confident enough in our programming abilities to implement
these correctly. Cannot propagate on negative decisions.

@@

\subsection{Instances}

?? Other studies use a random selection of 200 of the mesh instances, partly because some earlier
solvers find many of these instances extremely hard, and partly to avoid skewing a portfolio. We
want larger numbers of satisfiable instances, so we include all pattern / target pairs.

@@

?? Non-induced, compare to Glasgow, LAD, PathLAD, SND?, VF2, using tailored

@@

?? \cref{figure:value-ordering-heuristics}

@@

    \caption{Cumulative number of instances solved over time, using the five different
    value-ordering heuristics. The top three lines are very close together, with the weighted
    heuristic being worse for low runtimes and slightly better towards the end.}
    \label{figure:value-ordering-heuristics}
\end{figure}

@@

?? Results for DDS counting discrepancies both ways, maybe LDS too. Also look just at satisfiable
instances.

@@

?? \cref{figure:restarts}

?? breakdown by family

@@

    \caption{Instance by instance comparison of the basic algorithm with the tailored heuristic,
    versus weighted with restarts. Below, comparing instead with simple random with restarts.}
    \label{figure:scatter}
\end{figure}

@@

>>>>

% cref style
\crefname{algorithm}{Algorithm}{Algorithms}
\Crefname{algorithm}{Algorithm}{Algorithms}
\crefname{figure}{Figure}{Figures}
\Crefname{figure}{Figure}{Figures}
\crefname{table}{Table}{Tables}
\Crefname{table}{Table}{Tables}
\crefname{section}{Section}{Sections}
\Crefname{section}{Section}{Sections}

\title{Value Ordering, Discrepancies, and Restarts in Subgraph Solvers: \\ Some Half-Baked Mucking
Around}
\author{Ciaran McCreesh\thanks{This work was supported by the Engineering and Physical Sciences
    Research Council [grant number EP/026842/1]}\\ University of Glasgow, Glasgow, Scotland \\
    ciaran.mccreesh@glasgow.ac.uk}

@@

\begin{abstract}
    At least one modern subgraph isomorphism solver use reasonably good value-ordering heuristics to
    direct search, but these heuristics are not perfect, particularly for early choices. We
    investigate two mechanisms for offsetting this weakness: discrepancy searches, and restarts. Our
    results suggest that discrepancy search is too expensive to be effective, but that restarts with
    a random value-ordering heuristic beats the best tailored heuristic. We also introduce a
    slightly-random value-ordering heuristic with similar behaviour to the tailored heuristic.  When
    combined with restarts, this new heuristic is even more effective on satisfiable instances, and
    does not substantially affect performance on unsatisfiable instances.
\end{abstract}

@@

The current single strongest subgraph isomorphism solver uses ``highest degree first'' as a
value-ordering heuristic.  This is much better than not having a value-ordering heuristic at all,
but it is far from perfect, particularly for early branching choices. There are two reasons to
suspect we could do better.  Firstly, even if degree were a perfect source of information, many
graphs do not have a large degree distribution, and so we would still need to determine \emph{which}
vertex of highest degree would be best.  Secondly, experiments with a parallel search strategy which
explicitly steals early branching choices often gives strongly super-linear speedups on satisfiable
instances.

We could try to introduce tiebreaking mechanisms to further refine this heuristic. However, given
the extreme cost of making an incorrect branching choice early on in search, we believe it is more
fruitful to investigate alternatives to simple backtracking (depth-first) search. This paper
evaluates two such approaches: discrepancy search, and restarts.

?? Quick intro to discrepancy search. Quick intro to restarts, mentioning how it's important to
change something when restarting.

?? And concludes what?

@@

?? Started with the Glasgow algorithm, removed parts which did not contribute hugely to speedups,
and ended up with a simpler algorithm that performs as well in practice. In particular, large
graphs, no ILF, and not doing supplemental graphs of length 3. This is \cref{algorithm:sip},
ignoring the shaded parts, which are introduced in the context of restarts.

?? Briefly describe this, probably referring elsewhere as far as possible to save space.

\newcommand{\siplabel}[1]{\label{line:sip:#1}}
\newcommand{\siplineref}[1]{line~\ref{line:sip:#1}}
\begin{algorithm}[p]\DontPrintSemicolon\small\SetInd{0.1em}{1em}
    \begin{tikzpicture}[remember picture,overlay]
        \coordinate (restart1sc) at ($(pic cs:restart1s) + (0, 0.15)$);
        \coordinate (restart1ec) at ($(pic cs:restart1e) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart1sc) (restart1ec)] { };
        \coordinate (restart2sc) at ($(pic cs:restart2s) + (0, 0.15)$);
        \coordinate (restart2ec) at ($(pic cs:restart2e) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart2sc) (restart2ec)] { };
        \coordinate (restart3sc) at ($(pic cs:restart3s) + (0, 0.15)$);
        \coordinate (restart3ec) at ($(pic cs:restart3e) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart3sc) (restart3ec)] { };
        \coordinate (restart4sc) at ($(pic cs:restart4s) + (0, 0.15)$);
        \coordinate (restart4ec) at ($(pic cs:restart4e) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart4sc) (restart4ec)] { };
        \coordinate (restart5sc) at ($(pic cs:restart5s) + (0, 0.15)$);
        \coordinate (restart5ec) at ($(pic cs:restart5e) - (0, 0.10)$);
        \coordinate (restart5rc) at ($(pic cs:restart5r) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart5sc) (restart5ec) (restart5rc)] { };
        \coordinate (restart6sc) at ($(pic cs:restart6s) + (0, 0.15)$);
        \coordinate (restart6ec) at ($(pic cs:restart6e) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart6sc) (restart6ec)] { };
        \coordinate (restart7sc) at ($(pic cs:restart7s) + (0, 0.15)$);
        \coordinate (restart7ec) at ($(pic cs:restart7e) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart7sc) (restart7ec)] { };
        \coordinate (restart8sc) at ($(pic cs:restart8s) + (0.05, 0.15)$);
        \coordinate (restart8ec) at ($(pic cs:restart8e) + (0, 0.00)$);
        \node [fill=magma8, rounded corners, fit=(restart8sc) (restart8ec)] { };
    \end{tikzpicture}

@@

        \nl \tikzmark{restart2s}trigger nogoods containing only one clause\tikzmark{restart2e} \;

@@

\medskip

@@

    \nl $D_v \gets \textnormal{a domain from}~D~\textnormal{chosen by variable heuristic}$\siplabel{variableordering} \;

@@

    \nl \ForEach {$v' \in D_v~\textnormal{ordered by value heuristic}$\siplabel{valueordering}}{

@@

            \nl $S \gets \FuncSty{search}(\{ E \in D' : \left|E\right| > 1 \}, B \wedge (v \mapsto v'))$ \;

@@

                \nl post nogoods $\{B \wedge (v \mapsto c) \Rightarrow \bot : c \in C \}$\tikzmark{restart5r}\;

@@

        \nl \tikzmark{restart7s}trigger nogoods with watch $(v \mapsto v')$\tikzmark{restart7e} \;

@@

\caption{Subgraph isomorphism. Shaded code is to be ignored for the non-restarting algorithm.}
\label{algorithm:sip}

@@

For the variable-ordering heuristic on \siplineref{variableordering} of \cref{algorithm:sip}, we use
smallest domain first, tiebreaking on ?? degree. Variable ordering is not the focus of this
paper---as far as possible, we will leave it intact, so as not to affect the performance of the
algorithm on unsatisfiable instances.

For value ordering, which occurs on \siplineref{valueordering} of \cref{algorithm:sip}, we will
compare four strategies:

\paragraph{Random} We select a vertex dynamically at random (that is, we shuffle the vertices before
entering the $\KwSty{foreach}$ loop on \siplineref{valueordering} of \cref{algorithm:sip}). This is
\emph{usually} effectively equivalent to picking vertices in the order in which they appear in the
input files, although some instances from some benchmark suites are naturally in a special order
(which may or may not be a good order for searching).

\paragraph{Tailored} We select vertices from highest degree to lowest degree. This is done by
permuting the target graph at the top of search. This heuristic is the current state of the art.

\paragraph{Anti} We select vertices from lowest degree to highest degree. This heuristic is used as
a sanity check: we expect tailored to beat random, and random to beat anti.

\paragraph{Weighted} As with the random heuristic, we shuffle the vertices before we enter the
$\KwSty{foreach}$ loop. Instead of picking a vertex uniformly at random, we use a weight of
$2^{\left|\neighbourhood(v)\right|}$ when deciding whether to branch on vertex $v$ next.\footnote{We
cheat slightly to avoid range issues with random number generation: we give the vertex with highest
degree in the entire graph weight $2^{50}$, and then give lower degree vertices smaller powers,
giving every vertex weight at least 1.  This is a bit of hack but programming is hard.} This
heuristic heavily prefers vertices of high degree, similar to the tailored heuristic, but will give
the same chance of being selected next to two vertices of equal degree. This heuristic also
introduces an element of randomness, which will be important for restarts.

@@

?? Talk about how we're not dealing with a binary search tree, so we could either make it binary, or
count one discrepancy each time the degree changes, or count every ``against'' branch as being just
one discrepancy regardless.

@@

?? We use the Luby scheme, because everyone else does. We count the number of backtracks (that is,
when we reach the end of the main $\KwSty{foreach}$ loop) to decide when to restart. Following
established wisdom, we multiply each item in the Luby sequence by a magic constant. Preliminary
experiments demonstrated that this is useful, but we failed to determine a principled way of
selecting the constant's value, so we fell back on divine revelation and set it to 666. As with
everyone else, we are not entirely clear why we are doing this: it could be to allow the solver to
spend more time deep in search, it could be to reduce the overheads from repeatedly visiting inner
nodes, or it could be to keep the nogoods small.

?? We just use two watched literals. Could use increasing nogoods etc. We are not confident enough
in our programming abilities to implement these correctly, and we are not maintaining arc
consistency anyway. We also really do not want to propagate on negative decisions, because we do not
want to iterate over every removed value.

?? Talk a bit about how it fits into \cref{algorithm:sip}.

@@

?? Dual Xeon E5-2697A v4 CPUs, 512GBytes RAM, running Ubuntu 17.04, implemented in C++ and compiled
using GCC 6.3.0. Our code is heavily based upon ??.

?? Deterministic RNG.

?? Bitsets.

?? The 5,725 instances.

?? Other studies use a random selection of 200 of each of the mesh and images instances, partly
because some earlier solvers find many of these instances extremely hard, and partly to avoid
skewing a portfolio. We want larger numbers of satisfiable instances, so we include all pattern /
target pairs. This gives a total of 14,621 instances, ?? of which we know are satisfiable, and ??
unsatisfiable.

?? Note that an awful lot of these instances are very easy. We are trying to tackle the remaining
hard satisfiable instances.

@@

?? Establish a baseline. Compare to older Glasgow2, Glasgow3, LAD, PathLAD, SND if we can get it to
run on the larger set of instances, VF2, maybe SAT just to have something to cite showing how bad
SAT is at this. Just use tailored in this plot. Possibly just use the 5,725 instances here to avoid
pain with lousy solvers.

@@

In \cref{figure:value-ordering-heuristics} we show the cumulative number of instances solved over
time using the four different value-ordering heuristics. As expected, the tailored heuristic is much
better (?? give aggregate speedups for both all and just satisfiable instances) than random
ordering, which is in turn ?? times better than the anti-heuristic.

The performance of the weighted and tailored heuristics is very similar: the weighted heuristic has
slightly worse performance below a 100s timeout, and effectively equal performance above it. This
shows that we can introduce an element of randomness into the tailored value-ordering heuristic
without adversely affecting its performance.

@@

    \caption{Cumulative number of instances solved over time, using the four different
    value-ordering heuristics.}
    \label{figure:value-ordering-heuristics}
\end{figure}

@@

?? Code up LDS and DDS and the like properly, and plot them being disappointing here.

@@

In \cref{figure:restarts} we show the effects of adding restarts to the algorithm. With restarts,
the random variable-ordering heuristic comfortably beats the tailored strategy without restarts. In
other words, although having a good variable-ordering heuristic is beneficial, introducing
randomness into the search is better, if it is done alongside a mechanism to avoid heavy commitment
to any particular random choice.  However, the weighted heuristic together with restarts is better
still---that is, if we are introducing restarts, then it is better to add a small amount of
randomness to a tailored deterministic heuristic than it is to throw away the heuristic altogether.

@@

    \caption{Instance by instance comparison of the basic algorithm with the tailored heuristic,
    versus weighted + restarts. Below, comparing instead with simple random + restarts.}
    \label{figure:scatter}
\end{figure}

Cumulative plots only show aggregate behaviour. What happens on an instance by instance basis?  In
the top of \cref{figure:scatter} we scatter plot performance, showing satisfiable and unsatisfiable
instances using dots and crosses respectively. On the $x$-axis we show the tailored heuristic,
whilst on the $y$-axis we show weighted + restarts. Points on the outer axes represent instances
which timed out with one heuristic but not the other.

Nearly all of the unsatisfiable instances are on or slightly above the $x-y$ diagonal, showing that
their performance is either unchanged or reduced slightly. On the other hand, there are ?? large
numbers of satisfiable instances well below the diagonal line, indicating large speedups. Better
yet, there are only ?? satisfiable instances that are more than a factor of ?? times worse.

?? Figure out what's going on with the unsat instances not on the line. There are a few options.
Could be SDF picking something else, could be the hacky all-diff missing something, could be how
we're doing unit propagation, could be a horrible bug.

For contrast, the bottom plot of \cref{figure:scatter} compares the tailored heuristic with random +
restarts. The performance on unsatisfiable instances is the same. However, although better in
aggregate, we see many more satisfiable instances performing substantially worse. ?? Quantify this a
bit.

?? do a more detailed breakdown by family

?? Explain why this is so amazingly awesome and the most important idea ever to happen in the entire
history of science.

@@

?? This stuff is quite good.

?? Bigger CP picture
